Case-based Arabic Morphological Analysis

### 2010-12-16 trunk/tex/sarf.aclhlt.tex
in the controller

\section{Related Work }
\label{sec:related}

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is the non-deterministic 
composition of a controller and three
transducers that are the prefix, stem, and suffix tansducers. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

\transfalse
\begin{figure*}[tb]
\center{
\resizebox{1.7\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example composition of the prefix, stem and suffix transducers.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words
\transtrue

@@

Since $\Phi$ is now in an accept state in the stem transducer, it will
spawn $\Sigma$ that transitions into the root of the 
suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Sigma$ represents a valid analysis and can proceed.

@@

Finally, the case-based controller can interfere at any point in the 
analysis to make a decision like ignoring a composition that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce any output and therefore the composition
of the three transducers with the controller is a finite 
state Automaton (FSA).

@@

Morphological analysis partitions a given Arabic word
into a set of morphemes.
A morpheme is either an affix or a stem. 
We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of prefixes, suffixes and stems and assigns a category
to each morpheme. 
The relation between morphemes and categories is one 
to many.

A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of the string is a 
valid morphological analysis if  and only if we have
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant.
When users try to add their custom morphemes to these
lists they may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
is that of the second morpheme in the pair.

Formally we define an affix $f$ recursively to be either
an atomic affix $a$ or an atomic affix concatenated
to an affix and we write $f: a | a f$. 
Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Sarf only keeps atomic affixes and their categories.

\subsection{Affix linear transducer}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into a structure similar to those
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two structures are directed acyclic graphs. 
The absence of cycles reduces the 
traversal of the affix transducer to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix transducers encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the transducer corresponds to the last letter 
of a prefix in the prefix and suffix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsection{Stem linear transducer}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names. 
We obtained the set of proper names from online 
and biblical sources.

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is also linear where the accept
states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Non-deterministic composition of transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic composition 
of the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all linear and do not have cycles as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for brevity and pedagogical purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{An abstraction of the non-deterministic composition of 
    the prefix, stem and suffix transducers.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} below controls Sarf. 
It takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

@@

The arrows in Figure~\ref{f:composition} that transition
from prefix to stem, stem to suffix, and suffix to prefix
require compatibility constraints for 
the transition.

@@

\section{Islamic Literature Case Study}
\label{sec:islamic}

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on that Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an exhaustive mechanism of authentication
check as. This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

Our analysis accepts a book as input
and segments it into a vector of hadiths. 
We segment each individual hadith into two parts. 
The first part is the sanad and it
contains the chain of narrators who related the hadith
from Mohammad. 
The second part is the matn or the content
of the hadith. 
We are concerned to further explore the sanad and
we detect the chain of narrators and 
the relation that links each narrator to his ancestor and 
his predecessor in the chain. 
We also detect the full name of each narrator that is
usually composed of several proper names with connectors
in between.

We built our hadith case-based controller to target
the detection of proper names. 
This includes the task of finding compound names 
composed of several words such as \RL{`bd alr.hmn} often
appearing as ``run-on'' words.
The controller also targets words that mean narrate when
they appear in the neighborhood of multiple proper names.

@@

We illustrate the hadith case-based controller 
in Figure~\ref{f:hadith}. 
The transitions in the hadith controller are excited
by inputs from the Sarf morphological analyzer. 
A NAME label means that Sarf encountered a proper name.
An NRC label means that Sarf met a narrator connector such as
\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
\RL{'a_hbar} "told" or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
The NISBA label corresponds to an adjective that points to a person such 
as \RL{alma.sriy} "the Egyptian".

@@

The controller has four states that correspond to 
an abstract position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to the state NAME\_S whenever
NAME is reported by Sarf.
State NRC\_S means that the controller thinks it is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

>>>>

\section{Related work }
\label{sec:related}

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix deterministic tansducers. 
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

\transfalse
\begin{figure*}[tb]
\center{
\resizebox{1.7\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words
\transtrue

@@

Since $\Phi$ is now in an accept state in the stem transducer, it will
spawn $\Sigma$ that transitions into the root of the 
suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
$\Xi$ represents a valid analysis and can proceed.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct the analysis of one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce any external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
%The relation between morphemes and categories is one 
%to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of $s$ is a 
valid morphological analysis if and only if 
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair.

Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. We write $f: a | a f$. 
%Sarf only keeps atomic affixes and their categories.

\subsection{Affix transducers}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
This transducer is deterministic and thus 
its traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The affix transducers encode $L_{pp}$ and
$L_{xx}$.
An accept state in the transducer corresponds to the last letter 
of an affix in the affix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsection{Stem transducer}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names which we 
obtained from online and biblical sources.

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is deterministic and free of cycles where the 
accept states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Construction of Sarf from transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic construction of Sarf
from the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all deterministic and cycle free as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for pedagogical purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{An abstraction of the non-deterministic construction of  
    Sarf from the affix and stem transducers.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} 
takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

@@

%The arrows in Figure~\ref{f:composition} that transition
%from prefix to stem, stem to suffix, and suffix to prefix
%require compatibility constraints for 
%the transition.

@@

\section{Islamic literature case study}
\label{sec:islamic}

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on that Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an exhaustive mechanism of authentication
checks. This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper, we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

Our analysis accepts a book as input
and segments it into a vector of hadiths. 
We segment each individual hadith into two parts. 
The first part is the sanad and it
contains the chain of narrators who related the hadith
from Mohammad. 
The second part is the matn or the content
of the hadith. 
We further explore the sanad and
detect the chain of narrators and 
the relation that links each narrator to his ancestor and 
predecessor in the chain. 
We also detect the full name of each narrator that is
usually composed of several proper names with connectors
in between.

We built our hadith controller to target
the detection of proper names. 
This includes the task of finding compound names 
composed of several words such as \RL{`bd alr.hmn} often
appearing as ``run-on'' words.
The controller also targets words that mean ``narrate'' when
they appear in the neighborhood of multiple proper names.

@@

We illustrate the hadith controller 
in Figure~\ref{f:hadith}. 
The transitions are labeled by outputs generated from 
the Sarf transducers. 
A NAME label denotes a proper name.
An NRC label denotes a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
\RL{'a_hbar} ``told'' or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} ``the son of'' and is a name connector.
The NISBA label corresponds to an adjective that points to a person such 
as \RL{alma.sriy} ``the Egyptian''.

@@

The controller has four states that correspond to 
a position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to state NAME\_S on
NAME.
State NRC\_S means that the controller is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
